---
title: "Interpreting Experience Mediates Recruitment of Early Intonational Cues to Sentence Modality: An Eye Tracking Study"
shorttitle: "Interpreting Experience Mediates Recruitment of Early Intonational Cues to Sentence Modality: An Eye Tracking Study"
author:
  - name: Robert Esposito
    corresponding: true
    email: rme70@rutgers.edu
    affiliations:
      name: "Rutgers University"
      department: Department of Spanish and Portuguese
      address: 15 Seminary Pl.
      city: New Brunswick
      region: NJ
      country: USA
      postal-code: 08901
author-note:
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    # Example: This study was registered at X (Identifier Y).
    study-registration: ~
    # Acknowledge and cite data/materials to be shared.
    data-sharing: ~
    # Example: This article is based on data published in [Reference].
    # Example: This article is based on the dissertation completed by [citation].  
    related-report: ~
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: The authors have no conflicts of interest to disclose.
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: ~
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
number-sections: true
abstract: "Accurately parsing intonation is crucial for identifying sentence modality (i.e., declarative versus interrogative utterances) in Spanish. For example, *Mariano habla del tiempo* may variably be a statement ('Mariano talks about the weather') or a neutral information-seeking yes-no question ('Does Mariano talk about the weather?') depending on the intonation. Traditional analysis of Spanish intonation points to solely the final pitch movement as determining sentence modality [@tomas1944manual], but a recent gating experiment has demonstrated that L1 Spanish speakers can identify sentence modality by the first word [@face2007role]. Humans are believed to integrate linguistic information incrementally to make predictions about future content [@van2012prediction], and it follows that Spanish speakers would use early intonational cues to predict sentence modality instead of waiting for the final pitch movement, but there has yet to have been on-line measures to determine if sentence modality is predicted during real-time processing based on intonatioanl cues. This study uses eye-tracking to investigate L1 Spanish speakers' and L1 English L2 Spanish learners' sensitivity to early intonational information for identifying sentence modality. Participants completed a two-alternative forced choice task in which they identified if an orally-delivered utterance was a question or not. Results of this study provide insight into understanding if either or both populations are sensitive to non-vital intonational information to predict sentence type. The findings of this study shed light on the incremental processing of intonational information as real-time sentence comprehension unfolds and provides information about which cues are recruited to make predictions."
keywords: [keywords, go, here]
floatsintext: true
bibliography: "./lit/cog_interp_final.bib"
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf:
    documentmode: man
---

```{r}
#| include: false
# load libs
library(here)
```

When listening to speech, humans are exposed to an acoustic signal that is seldom encoded in writing: fundamental frequency (F0). F0, modulated by how fast the vocal folds vibrate, is interpreted in human language as discrete pitch targets (high or low) and organized at the utterance-level to provide syntactic, pragmatic, and paralinguistic information into what we nominate intonation [@ladd2008intonational]. Both native speakers (L1 speakers) and second language learners (L2 speakers) must accurately produce or perceive intonation to prevent communicative breakdowns. For example, distinguishing a sarcastic exclamation and a genuinely complimentary one in English (e.g., "How smart she is!") requires correct parsing of intonation, as the utterance is identical at both the lexical (word) and syntactic (word-order) levels [@zhou2019detecting].

Even more quotidian, Spanish declaratives and yes-no questions are similarly distinguished only at the level of intonation: *Mariano habla del tiempo* "Mariano talks about the weather" may variably be a statement or information-seeking (i.e., neutral) yes-no question depending solely on intonation. Traditionally, only the final F0 movement of the utterance is analyzed as providing the disambiguating cue. However, earlier cues have been identified [@tomas1944manual; @quilis1993filologia; @verdugo2005aproximacion; @prieto2004search]. @face2007role determined in an off-line two-alternative forced choice task gating experiment that L1 Spain Spanish speakers can disambiguate sentence type as early as the first word, which encompasses the first F0 peak. While the final F0 movement strongly influenced participants' judgement, it appears that L1 Spanish speakers can use early intonational cues to determine sentence modality before the final movement is reached.

As of yet, there has been no investigation to determine if the earlier cues are recruited as sentence processing unfolds to predict sentence type. Theoretical models of brain function suggest that the brain makes continuous predictions about future events based on Bayesian inferences [@bar2007proactive; @lee2003hierarchical; @van2012prediction; @altmann1999incremental; @gruter2021prediction]. It is expected that L1 Spanish and "proficient" (that is, native-like) L2 Spanish speakers would recruit intonational information as sentence processing unfolds to predict sentence modality. 

Furthermore, since these Bayesian-based models of the brain rely prior experience, it is predicted that experiential-factors will have a strong influence on the outcome, such that those who have more reliable success in predicting sentence type based on the identified cues will pay more attention to those same cues in the future to make the same predictions [@roettger2019evidential]. Interpreters are trained to make use of prediction during interpreting [@dong2020attentional], and the same skills transfer to non-interpreting situations [@lozano2021interpreting]. Therefore, it is predicted that Spanish-English interpreters who more often interpret questions from Spanish to English, as opposed to English to Spanish, will more often recruit early intonational cues to make predictions on sentence modality.

The present study aims to determine if prediction of sentence modality in the L1 English L2 Spanish late learner interpreters is dependent on the frequency with which they must do so. Specifically, it is predicted that interpreters who work in a Spanish-speaking court system who must interpret questions more often from Spanish to English, as opposed to those who work in an English-speaking law system, will be more sensitive to early cues to sentence modality during on-line processing. The results will shed light on the weighting of early intonational cues compared to the final F0 movement in their usefulness to predicting sentence modality during real-time processing [@roettger2019evidential], as well as the impact of experiential factors, like interpreting experience.

## Early Cues to Sentence Modality in Spanish

Broadly speaking, intonation can be used to signal speech act marking, information status, belief status, and politeness, among many other communicative functions [@prieto2015intonational; @best2019diversity; @arvaniti2020autosegmental; @cruttenden1997intonation; @cutler1997prosody; @dahan2015prosody; @gussenhoven2004phonology; @ladd2008intonational]. Relevant to this study, intonation may also encode sentence modality: @hirst1998survey found that approximately 70% of a sample of nearly 250 languages used the final pitch movement to signal interrogativity. Spanish is an example of such a language, in which an utterance e.g., 'Mariano habla del tiempo' *Mariano talks about the weather* may variably be a statement or question, differentiated only by intonation. Importantly, @hirst1998survey categorized languages using only the final F0 movement, and this has been the traditional analysis for Spanish as well [@tomas1974manual; @hualde2008intonation].

Although the final F0 movement is seemingly the most salient in Spanish for determining sentence modality, it serving as the only cue has been unequivocally rejected by @face2005f0['s] off-line gating experiment after previous investigations into earlier cues [@prieto2004search; @quilis1993filologia; @verdugo2005aproximacion]. He found that L1 Spanish speakers can accurately determine sentence modality by the scaling (i.e., height) of the first F0 peak and the presence or absence of a medial pitch accent. Although speakers are clearly able to recruit the early intonational cues to predict sentence type off-line, it remains unclear if they recruit the same information during on-line sentence processing.

## Using Intonation to Predict

It seems unlikely that Spanish listeners should wait until the very end of an utterance to determine sentence modality. Humans constantly make predictions [@bar2007proactive; @lee2003hierarchical], and it follows that we also anticipate linguistic information [@van2012prediction]. Such a hypothesis has been supported across linguistic domains [e.g., @altmann1999incremental]. Prosody and intonation have traditionally received little attention [@cutler1997prosody], but the extant research has demonstrated that listeners integrate intonational information as sentence processing unfolds to make predictions of upcoming content [@esteve2020empathy; @grodner2010some; @kamide2003time; @crocker2010computational; @levy2008expectation]. Based on this evidence, it would be expected that L1 Spanish speakers do not wait until the final F0 movement to determine sentence modality, but instead integrate early intonational information to make predictions.

Traditionally, only the last F0 movement is pointed to as the meaningful difference between the two sentence types [@tomas1974manual; @hualde2008intonation], but other non-final intonational cues have been identified [@sensui1995perception; @sensui2003pilot; @face2004intonation; @face2005f0]. @face2005f0['s] off-line gating experiment unequivocally rejected the final F0 movement as the only cue for sentence modality: L1 Spain Spanish speakers can accurately determine sentence modality by the scaling (i.e., height) of the first F0 peak and the presence or absence of a medial pitch accent. In a follow-up off-line gating experiment [@face2007role], he found that these early cues hold dramatically less weight than the final F0 movement, which seems to "override" the sentence type interpretation regardless of which cues occurs to the left.

Although cues may be available, it is not given that a perceiver will integrate the cue in real-time processing to make predictions; that is, not all cues are weighted equally. @roettger2019evidential attempts to address this issue by introducing a computational model that proposes that the predictive value of an intonational cue is dependent on the prior reliability of the cue-to-function mapping and the likelihood of the mapping given the listener's most recent experiences. In a later series of mouse-tracking studies, @roettger2021positional found that L1 German speakers do not use the presence of absence of a prenuclear pitch accent to predict the discourse status of an upcoming referent, but some L1 American English speakers, who have the same cue with the same predictive power available, do. This research puts doubt on the claim that L1 Spanish speakers recruit early intonational cues, as opposed to waiting for the final F0 movement.

## Experiential factors

Based on @roettger2021positional['s] model, it's expected that one's experiences will influence the predictive cue weights, such as interpreting experience. Interpreters are seen as expert predictors [@dong2020attentional; @setton2005pointing], and they extend language prediction to non-interpreting linguistic contexts [@lozano2023interpreting; @lozano2021interpreting]. For example, in an eye-tracking study, @lozano2020slowly compared L1 English L2 Spanish learners with and without interpreting experience in their ability to anticipate a word's suffix based on the first syllable's prosodic patterns and found that those with interpreting experience demonstrated greater prediction than those without.

However, the phenomenon under investigation here is sensitive to non-linguistic factors. That is, interpreters working in a Spanish-speaking law system most likely interpret questions from lawyers from Spanish to English (e.g., during an examination), whereas those working in an English-speaking law system do not interpret questions as frequently. This would be an interesting situation to examine @roettger2021positional['s] model, which would predict that those who interpret questions from Spanish to English, as opposed to English to Spanish, would more readily recruit early intonational cues to predict sentence modality.

## Research Questions

Early intonational cues, such as first F0 peak and the presence or absence of a medial pitch accent[@tomas1944manual; @quilis1993filologia; @verdugo2005aproximacion; @prieto2004search; @face2005f0; @face2007role], allow for prediction of sentence modality to take place, but it is not yet clear if, as predicted by Bayesian models of the brain [@bar2007proactive; @lee2003hierarchical; @van2012prediction; @altmann1999incremental; @gruter2021prediction], L1 Spanish monolinguals or L1 Spanish L2 English late learners recruit such cues for prediction of sentence modality during on-line sentence processing. Experiential factors may also influence the degree to which the early intonational cues are recruited, such that interpreters with more experience interpreting questions from Spanish to English will show greater recruitment of earlier intonational cues.

The following research questions are experimental in nature:

1. Do L1 Spanish monolinguals recruit early intonational cues (i.e., first F0 speak and presence/absence of medial pitch accent) to predict sentence modality?

2. Do L1 English L2 Spanish late learners recruit early intonational cues (i.e., first F0 speak and presence/absence of medial pitch accent) to predict sentence modality?

3. Is the recruitment of the identified early intonational cues in L1 English L2 Spansh late learner interpreters dependent on experience interpreting questions from Spanish to English?

# Method

## Participants

Thirty L1 Spanish monolinguals, 30 L1 English L2 Spanish late learner interpreters in central Spain, and 30 L1 English L2 Spanish late learner interpreters in Gibraltar were recruited. The L1 Spanish monolinguals grew up in central Spain and have not spent more than 3 months outside of their city of origin. The L1 English participants were raised in a monolingual English environment in England and began the endeavor of learning Spanish after 12 years old. 

## Materials

Aside from the experimental task, participants also completed a language background questionnaire. The survey collected information about participants' age of acquisition, cumulative language use, and current language use. For interpreters, it asked about how often they interpreted questions from Spanish to English. L2 participants were seen to have acquired Spanish primarily in the classroom, and all had experience living in Spain.

Using the visual-world eye-tracking paradigm, participants were tested to determine if they were able to accurately determine whether a sentence was a declarative or interrogative before the final syllable. There were 60 critical trials. 30 were declarative utterances and 30 were interrogative utterances.

All utterances consisted of an overt subject, a verb, and an object e.g., 'Mariano habla del tiempo' *Mariano talks about the weather*. The final word of the utterance always had penultimate stress so that the final pitch movement was not truncated with the nuclear pitch accent.

The utterances were all produced by a 25-year-old female L1 Spanish monolingual speaker from Madrid. In total, she produced 30 pairs of identical utterances for a total of 60 utterances: half of the utterances were produced as declaratives, and the other half were produced as interrogatives. Acoustic analysis revealed that the first pitch peak of declarative utterances were always lower than the first pitch peak of the interrogative utterances. The declarative utterances were always produced with a medial pitch accent (e.g., on 'habla' in 'Mariano habla del tiempo'), whereas this pitch accent was absent in interrogative utterances. As expected, declarative utterances ended with a low boundary tone, whereas interrogative utterances ended with a high boundary tone.

The visual display accompanying the auditory stimuli had two areas of interest depicting identical images of two people. One of the people had a speech bubble with either '.' or '?'. An example trial is depicted in @fig-example-trial.

```{r}
#| label: fig-example-trial
#| fig-cap: Example of one trial.

knitr::include_graphics(
  here("assets", "example_trial.bmp")
  )
```

## Procedure

Prior to the eye-tracking task, all participants completed the LexTALE-ESP and the language background questionnaire. Once completed, the participant's dominant eye was determined by applying the Miles test [@miles1929ocular]. Only eye movements from the dominant eye were registered. Participants were seated in front of an Eyelink Pro 1000 eye-tracking system (approximately 80 cm). Participants wore headphones to hear audio stimuli. Participants were told that they would hear a number of sentences and must determine if it was a 'statement', which corresponded to the image with a '.', or a 'yes-no question', which corresponded to the image with a '?'. The placement of the visual stimuli were counterbalanced. Participants were asked to press either the left or right shift key that corresponded with the image on the left or right side of the screen that matched the utterance's sentence type, and they could respond at any point during the utterance.

At trial start, participants were exposed to a fixation cross for 500ms, after which the two images appeared for 200ms. At 700ms into the trial, the audio stimuli began to play. Once the audio finished, participants could respond using the shift keys. Once they responded, a grey box appeared on the screen in one of the four corners. After they looked at the box for 500ms, the next trial began.

Participants completed 4 practice trials before the experiment to familiarize themselves with the images and task. Participants were then exposed to a total of 90 (30 experimental and 60 filler) trials. The task took approximately 15 minutes to complete.

# Results

# Discussion

## Limitations and Future Directions

## Conclusion

# RANDOM SNIPPETS, DELETE ME LATER 

With specific regards to intonational cues, @foltz2021using in an eye-tracking study found that L1 German L2 English intermediate-to-advanced learners [B2 or above using CEFR levels; @coucilofeurope2001cefr] were able to predict a referent based on a contrastive pitch accent in their L1, but showed delayed processing for the same cue-to-function mapping in their L2. This is in contrast to @klassen2015second, who found that intermediate-to-advanced Spanish-English bilinguals *did* engage in predictive processing for a similar cue-to-function mapping between languages.

@marasco2020you Gating experiment a la Face 2007. L2 = native speaker for reaction times at gate 1. Three gates. The first right after first pitch peak. Statistics only completed for first date; third gate's RTs are reported, but statistics are not completed on it.

Importantly, they attribute these differing results to timing: where @klassen2015second separated two phrases of an instruction with 700 ms of silence, the target noun in @foltz2021using immediately followed the cue-bearing adjective. Thus, it may be expected that if given enough time (i.e., a longer sentence length), L2 Spanish speakers may integrate early intonational cues.

An open question is *which* information perceivers recruit to make predictions about *what* content. 

As such, although it is clear that L1 Spanish speakers *can* recruit early intonational cues to accurately predict sentence modality [@face2007role], it is yet to be seen if they *do* recruit the early cues during real-time sentence processing.

For example, @esteve2020empathy in an eye-tracking experiment found that L1 French speakers can recruit intonational information to accurately disambiguate cognates. @ito2008anticipatory investigated L1 English speakers' perception of the felicitous and infelicitous use of the L+H* pitch accent, associated with contrastive focus, in one of their three eye-tracking experiments reported. In experiment 2, they found that listeners had earlier target fixations when presented with the contrastive pitch accent and demonstrated a garden path effect when the contrastive pitch accent was used infelicitously. These studies indicate that intonational information *can* recruited during real-time sentence processing to predict upcoming information.

In an off-line gating experiment, @face2007role determined that L1 Spanish speakers are capable of recruiting early intonational 

but it remains unclear if speakers do recruit the relevant early intonational information during real-time sentence processing.

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::

# This Section Is an Appendix {#apx-a}


# Another Appendix {#apx-b}